% This is part of the book TeX for the Impatient.
% Copyright (C) 2003 Paul W. Abrahams, Kathryn A. Hargreaves, Karl Berry.
% See file fdl.tex for copying conditions.

\input macros
\chapter{Using this book}

\chapterdef{usebook}

This book is a do-it-yourself guide and handbook for \TeX. 
Here in this section we tell you how to use the book to maximum advantage.

We recommend that you first either read or skim in sequence Sections
\chapternum{usebook} through \chapternum{examples},
which tell you what you need to know in order to get started using \TeX.
If you've already had experience using \TeX, it will still be helpful
to know what kinds of information are in these sections of the book.
Sections~\chapternum{concepts}--\chapternum{tips}, which
occupy most of
the rest of the book, are designed to be accessed randomly.
Nevertheless, if you're the kind of person who likes to read reference manuals,
you'll find that it \emph{is} possible to proceed sequentially if you're
willing to take a lot of detours at first.

In \chapterref{usingtex}, ``Using \TeX'',
we explain how to produce a \TeX\ document from a
\TeX{} input file.
We also describe the conventions for preparing that input file,
explain a little about how \TeX{} works, and tell you about additional
resources that are available.
Reading this section will help you understand
the examples in the next section.

\chapterref{examples}, ``Examples'',
contains a sequence of 
examples that illustrate the use of \TeX.
Each example consists of a page of output
together with the input that we used to create it.
These examples will orient you and help you 
locate the more detailed material that you'll need as you go.  
By seeing which commands are used in the input, you'll know
where to look for more detailed information on how to achieve
the effects shown in the output.
The
examples can also serve as models for simple documents, although we must
caution you that 
because we've tried to pack a variety of \TeX\ commands into a small
number of pages,  
the examples are not necessarily illustrations of good or complete 
document~design.

As you read the explanation of a command, you may encounter
some unfamiliar technical terms.
In \chapterref{concepts}, ``Concepts'',
we define and explain these terms.
We also discuss other topics that aren't covered
elsewhere in the book.
The inside back cover of the book contains a list of all the
concepts and the pages where they are described.
We suggest that you make a copy of this list and keep it nearby
so that you'll be able to identify and look up an unfamiliar
concept immediately.

\TeX's commands are its primary vocabulary,
and the largest part of this book is
devoted to explaining them.  In Sections~\chapternum{paras}
through~\chapternum{general} we describe the commands.
You'll find general information about the command descriptions
on \xrefpg{cmddesc}.
The command descriptions are arranged
functionally, rather like a thesaurus, so if you know what you want to
do but you don't know which command does it for you, you can use the
table of contents to guide you to the right group of commands.
Commands that we think are both particularly useful and easy to understand
are indicated with a pointing~hand~(\hand).

\chapterref{capsule}, ``Capsule summary of commands'', is a
specialized index that complements the more complete descriptions
in Sections~\chapternum{paras}--\chapternum{general}.
It lists \TeX's commands 
alphabetically, with a brief explanation of each command
and a reference to the page
where it is described more completely.  The capsule summary
will help you when you just want a quick reminder of what a command
does.

\TeX\ is a complex program that occasionally works its will in
mysterious ways.
In \chapterref{tips}, ``Tips and techniques'',
we provide advice on solving a variety of specific 
problems that you may encounter from time to time.
And if you're stumped by
\TeX's error messages, you'll find succor in \chapterref{errors},
``Making sense of error messages''.

The gray tabs on the side of the book will help you locate parts of the
book quickly.  They divide the book into the following major parts:
\olist
\li general explanations and examples
\li concepts
\li descriptions of commands (five shorter tabs)
\li advice, error messages, and the |eplain.tex| macros
\li capsule summary of commands
\li index
\endolist

In many places we have provided page references to
^{\texbook} (see \xrefpg{resources} for a citation).
These references apply to the seventeenth edition of \texbook.
For other editions, some references may be off by a page or two.


\section Syntactic conventions

In any book about preparing input for a computer,
it's necessary to indicate clearly the literal characters that should be typed
and to distinguish those characters from the explanatory text.
We use the Computer Modern typewriter font for {\tt literal input
like this}, and also for the names of \TeX{} commands.
When there's any possibility of confusion, we enclose \TeX\
input in single quotation marks, `{\tt like this}'.
However, we occasionally use parentheses when we're indicating single
characters such as (|`|) (you can see why).

For the sake of your eyes we usually just put spaces 
where you should put spaces. In some places where
we need to emphasize the space, however,
we use a `\visiblespace' character
{\catcode `\ =\other\pix^^| |}%
to indicate it.
Naturally enough, this character is called a \emph{visible space}.
\pix^^{spaces//visible}


\section Descriptions of the commands

\xrdef{cmddesc}
Sections~\chapternum{paras}--\chapternum{general} contain
a description of what nearly every \TeX\ command does.  ^^{commands}
Both the primitive commands ^^{primitive//command}
and those of ^{\plainTeX} are covered.
The primitive commands are those built into the \TeX\ computer program, while
the \plainTeX{} commands are defined in a standard file of 
auxiliary definitions (see \xref\plainTeX). 
The only commands we've omitted are those that are used purely locally
in the definition of \plainTeX\ (\knuth{Appendix~B}).
The commands are organized as follows:
\ulist\compact
\li ``Commands for composing paragraphs'', \chapterref{paras},
deal with characters, words, lines, and entire paragraphs.
\li ``Commands for composing pages'', \chapterref{pages},
deal with pages, their components, and the output routine.
\li ``Commands for horizontal and vertical modes'', \chapterref{hvmodes},
have corresponding or identical
forms for both horizontal modes (paragraphs and hboxes) and vertical
modes (pages and vboxes).
These commands provide boxes, spaces, rules, leaders,
and alignments.
\li ``Commands for composing math formulas'', \chapterref{math},
provide capabilities for constructing math formulas.
\li ``Commands for general operations'', \chapterref{general},
provide 
\TeX's programming features and
everything else that doesn't fit into any of the other sections.
\endulist
You should think of these categories as being suggestive rather than
rigorous, because the commands don't really fit neatly into these
(or any other) categories.
 
Within each section, the descriptions of the commands are organized
by function.  When several commands are closely related, they are described as
a group; otherwise, each command has its own explanation.
The description of each command
includes one or more examples and the output
produced by each example when examples are appropriate (for
some commands they aren't).
When you are looking at a subsection containing functionally related
commands, be sure to check the end of a subsection for a ``see also'' 
item that refers you to related commands that are described elsewhere.
 
Some commands are closely related to certain concepts.
For instance, the |\halign| and |\valign| commands are related to
``alignment'', the |\def|
command is related to ``macro'',
and the |\hbox| and |\vbox| commands are related to ``box''.
In these cases we've usually given a bare-bones des\-crip\-tion of the
commands themselves and explained  the underlying ideas
in the concept. 

The examples associated with the commands have been typeset with
^|\parindent|, the paragraph indentation, set to zero so that
paragraphs are normally unindented.
This convention makes the examples easier to read.
In those examples where the paragraph indentation is essential,
we've set it explicitly to a nonzero value.

The pointing hand in front of a command or a group of commands indicates
that we judged this command or group of commands to be particularly useful
and easy to understand.

Many commands expect ^{arguments} of one kind or another
(\xref{arg1}).  The arguments of
a command give \TeX\ additional information that it needs in order to
carry out the command.  Each argument is indicated by an italicized
term in angle brackets that indicates what kind of argument it~is:
 
\display{%
\halign{\<#>\quad&#\hfil\cr
argument&a single token or some text enclosed in braces\cr
charcode&a character code, i.e., an integer between $0$ and $255$\cr
dimen&a dimension, i.e., a length\cr
glue&glue (with optional stretch and shrink)\cr
number&an optionally signed integer (whole number)\cr
register&a register number between $0$ and $255$\cr
}}
^^{\<dimen>}
^^{\<argument>}
^^{\<charcode>}
^^{\<glue>}
^^{\<number>}
^^{\<register>}

\noindent
All of these terms are explained in more detail in \chapterref{concepts}.
In addition, we sometimes use terms such as \<token list> that are either
self-explanatory or explained in the description of the command.
Some commands have special formats that require either braces or 
particular words.
These are set in the same bold font that we use
for the command headings.

Some commands are parameters (\xref{introparms}) or table entries. 
^^{parameters//as commands}
This is indicated in the command's listing.
You can either use a parameter as an argument or assign a value to it.
The same holds for table entries.
We use the term ``parameter'' to refer to entities such as |\pageno|
that are actually registers but behave just like parameters.
^^{registers//parameters as}


\endchapter
\byebye
